```{r}
install.packages("tidyverse")
install.packages("naniar")
install.packages("leaflet")
library(tidyverse)
library(dplyr)
library(naniar)
library(lubridate)
library(leaflet)
library(ggplot2)
```

```{r}
file_path <- "/Users/Jeevikamenghwani/Documents/chicago_violations.csv"
chicago_violations <- read.csv(file_path)
View(chicago_violations)
str(chicago_violations) 
```
#The `str()` function in R offers insights into the structure of a dataframe. In this instance, it reveals that the dataframe called `chicago_violation` comprises 784,225 rows and 22 variables.

#Here's a breakdown of the dataframe's structure:

#- It consists of 784,225 observations (rows) and 22 variables (columns).
#- Each row represents a distinct violation record.
#- The dataframe encompasses a variety of data types, including character, integer, numeric, and factor.
#- The columns contain information pertaining to violations, such as addresses, dates, dispositions, fines, and geographical coordinates.

#This summary provides a comprehensive overview of the dataframe's composition and the nature of the data it holds.
```{r}
chicago_violation_of_ward_15 <- filter(chicago_violations, WARD == 15)
View(chicago_violation_of_ward_15)
str(chicago_violation_of_ward_15)
```
# It consists of 16383 observations (rows) and 22 variables (columns).
# Intresting facts about 15th watrd
#1. 15th ward,  covers parts of the West Side, including Brighton Park, Gage Park, New City (Canaryville), West Englewood and Back of the Yards.
#2. Current Alderman :  Raymond A. Lopez
#3. The first great manufacturing industry in Chicago was located in the 15 Ward in 1857, when the North Chicago rolling mills were established
```{r}
any_na <- any(is.na(chicago_violation_of_ward_15))
print(paste("are there any NA?",any_na))
total_na_count <- sum(is.na(chicago_violation_of_ward_15))
 print(paste("total na count is:",total_na_count))

complete_cases <- complete.cases(chicago_violation_of_ward_15)
 num_complete_cases <- sum(complete_cases)
 print(paste("total complete cases are:",num_complete_cases))
 total_rows <- nrow(chicago_violation_of_ward_15)
 percentage_complete_cases <- (num_complete_cases/total_rows)*100
 print(paste("perecentage of complete cases:",percentage_complete_cases))
```


```{r}
chicago_violation_of_ward_15 [chicago_violation_of_ward_15 == ""] <-NA
View(chicago_violation_of_ward_15)
 complete_cases <- complete.cases(chicago_violation_of_ward_15)
 num_complete_cases <- sum(complete_cases)
 print(paste("number of complete cases after filling blanks:",num_complete_cases))
 total_rows <- nrow(chicago_violation_of_ward_15)
 percentage_complete_cases <- (num_complete_cases/total_rows)*100
print(paste("percentage of complete cases after filling blanks:",percentage_complete_cases))
 total_na_count <- sum(is.na(chicago_violation_of_ward_15))
 print(paste("total na count after filling blanks:",total_na_count))
```

```{r}
missing_values <- miss_var_summary(chicago_violation_of_ward_15)
print (missing_values)
```

```{r}
str(chicago_violation_of_ward_15)
```
# the variables that has date is - HEARING.DATE, LAST.MODIFIED.DATE,VIOLATION.DATE. 
# R Views them as charecter 

```{r}
chicago_violation_of_ward_15$HEARING.DATE <-as.Date(chicago_violation_of_ward_15$HEARING.DATE, format ="%m/%d/%Y" )
chicago_violation_of_ward_15$LAST.MODIFIED.DATE <- mdy_hms(chicago_violation_of_ward_15$LAST.MODIFIED.DATE)
chicago_violation_of_ward_15$VIOLATION.DATE <- mdy_hms(chicago_violation_of_ward_15$VIOLATION.DATE)
str(chicago_violation_of_ward_15)
```


```{r}
chicago_violation_of_ward_15$HEARING.DATE <- as.POSIXct(chicago_violation_of_ward_15$HEARING.DATE)
chicago_violation_of_ward_15$VIOLATION.DATE <- as.POSIXct(chicago_violation_of_ward_15$VIOLATION.DATE)
chicago_violation_of_ward_15$CityDelay <- difftime(chicago_violation_of_ward_15$HEARING.DATE, chicago_violation_of_ward_15$VIOLATION.DATE, units = "days")
print(chicago_violation_of_ward_15)
str(chicago_violation_of_ward_15)
```


```{r}
```
# 

```{r}
chicago_violation_20_oct <- filter(chicago_violation_of_ward_15, day(VIOLATION.DATE) == 20 & month(VIOLATION.DATE) == 10) 
total_rows1 <- nrow(chicago_violation_20_oct)
print(paste("Total number of ordinance violations issued on October 20th:", total_rows1))


Most_common_CaseDisposition <- table(chicago_violation_20_oct$CASE.DISPOSITION) %>%   sort(Most_common_CaseDisposition, decreasing = TRUE)
Most_common_CaseDisposition <- names(Most_common_CaseDisposition)[1]
print(paste("most common case disposition:", Most_common_CaseDisposition))

```
# total number of ordinance violations are 44 
# most common case disposition is liable 
```{r}
cor(chicago_violation_of_ward_15$ADMIN.COSTS,chicago_violation_of_ward_15$IMPOSED.FINE)
Most_common_StreetType <- table(chicago_violation_of_ward_15$STREET.TYPE)%>%   sort(decreasing = TRUE)
Most_common_StreetType <- names(Most_common_StreetType)[1]
print(paste("Most common street type:", Most_common_StreetType))

Unique_violation_description <- length(unique(chicago_violation_of_ward_15$VIOLATION.DESCRIPTION))
Unique_violation_code<- length(unique(chicago_violation_of_ward_15$VIOLATION.CODE))
print(paste("Unique violation description:", Unique_violation_description))
print(paste("Unique_violation_code:",Unique_violation_code))
```
#The correlation coefficient between imposed fine and administrative costs is approximately 0.2242.indicating a weak positive correlation.This means that as one variable (imposed fine) increases, the other variable (admin costs) also tends to increase slightly, and vice versa. However, the strength of this correlation is weak, meaning that the relationship between imposed fine and admin costs is not very strong.
```{r}
Avg_Imposed_fine <-chicago_violation_of_ward_15 %>%
  mutate(year= year(HEARING.DATE)) %>%
  group_by(year) %>%
  summarise(AVG_IMPOSED_FINE = mean(IMPOSED.FINE , na.rm = TRUE))
Avg_Imposed_fine

```
# 2024 has the lowest average impoosed fines 
# therefore i expored the data more in the below code 
```{r}
Hearing_year <-chicago_violation_of_ward_15 %>%
  mutate(year= year(HEARING.DATE)) %>%
  group_by(year)
Chicago_violation_2024<- filter(Hearing_year,year == 2024)
casedisposition_2024 <- table(Chicago_violation_2024$CASE.DISPOSITION)
casedisposition_2024
```
# as imposed fines are based on case disposition and the major case disposition in 2024 is continuance meaning a legal decision to postpone court proceedings to a later date.Due to sevral reasons like - requests for more time to prepare the case, the unavailability of a party or witness, or the need for additional information or evidence.Cause of which there is no verdict and fines are not imposed. 


```{r}
chicago_violation_of_ward_15_2 <- chicago_violation_of_ward_15 [,-c(1,2)]
```

```{r}
chicago_violation_of_ward_15_2 <- chicago_violation_of_ward_15_2 %>%
  mutate(quarter= quarter(VIOLATION.DATE))

View(chicago_violation_of_ward_15_2)

chicago_violation_of_ward_15_2 <- chicago_violation_of_ward_15_2 %>%
  mutate(season = case_when(
    quarter == 1 ~ "Winter",
    quarter == 2 ~ "Spring",
    quarter == 3 ~ "Summer",
    quarter == 4 ~ "Fall",
    TRUE ~ as.character(quarter)
  )) %>%
  select(-quarter)
```

```{r}
ggplot(chicago_violation_of_ward_15_2, aes (x=season)) + geom_bar (fill= "red")+
labs(title =  "Building Violations Counts by Season")
```
# summer has the most violation, as the weather conditions support to do more of construction works like repair , replace or build something new 
```{r}
most_common_case_deposition <- table(chicago_violation_of_ward_15_2$CASE.DISPOSITION) %>% 
  sort(decreasing = TRUE)
most_common_case_deposition
chicago_violation_of_ward_15_3 <- filter(chicago_violation_of_ward_15_2, CASE.DISPOSITION %in% c("Liable","Non-Suit","Default","Not Liable","Continuance"))
View(chicago_violation_of_ward_15_3)  
Total_no_row2 <- nrow(chicago_violation_of_ward_15_3)
print(paste("Total number of rows:",Total_no_row2))

ggplot(chicago_violation_of_ward_15_3, aes(x = as.numeric(CityDelay, units = "days"), fill = ..count..)) +
  geom_histogram(binwidth = 1000, color = "black") +
  facet_wrap(~ `CASE.DISPOSITION`) +
  scale_fill_gradient("Count", low = "pink", high = "green") +
  labs(title = "Distribution of City Delay by Case Disposition",
       x = "City Delay (DAYS)",
       y = "Frequency") +
  theme_minimal()
```

```{r}
Avg_Imposed_fine_2 <-chicago_violation_of_ward_15_3 %>%
  group_by(CASE.DISPOSITION) %>%
  summarise(AVG_IMPOSED_FINE = mean(IMPOSED.FINE , na.rm = TRUE))

Avg_Imposed_fine_2$CASE.DISPOSITION <- reorder(Avg_Imposed_fine_2$CASE.DISPOSITION, Avg_Imposed_fine_2$AVG_IMPOSED_FINE, FUN = mean, na.rm = TRUE)

ggplot(Avg_Imposed_fine_2, aes (x=CASE.DISPOSITION, y= AVG_IMPOSED_FINE)) + geom_bar (stat = "identity", fill = "yellow", color = "green") +
  labs(title =  "imposed fines",
       x = "CASE DISPOSITION",
       y = "AVERAGE IMPOSED FINE") 


```
#This plot illustrates the average imposed fine associated with each type of case disposition. Upon analyzing the plot, several insights can be drawn:

#1. Cases labeled as "Default" exhibit the highest average imposed fines. This could be attributed to these cases being closed due to inactivity or a failure to proceed in a timely manner, resulting in a default judgment in favor of the plaintiff.
#2. "Liable" cases rank second-highest in terms of average imposed fines. This suggests that offenders in these cases are held accountable for the damage and are required to pay fines accordingly.
#3. "Not Liable" cases appear towards the lower end of the spectrum, indicating that these offenders are not held responsible for the damage and thus incur lower fines.
#4. "Continuance" cases demonstrate relatively lower average fines. This could be due to these cases being postponed or delayed, resulting in less severe consequences at the current stage.
#5. "Non-Suit" cases exhibit the lowest average imposed fines. This indicates that these cases have been dismissed, possibly due to insufficient evidence or other legal factors.

#Overall, the dominance of "Default" cases in terms of high fines suggests that many offenders may not actively engage in legal proceedings, leading to default judgments and subsequent fines. This trend also highlights the challenges in distinguishing between liable and non-liable offenders, potentially indicating a tendency for individuals to opt for paying fines without contesting the charges.

```{r}
ggplot(chicago_violation_of_ward_15_3, aes(x =ADMIN.COSTS, fill = CASE.DISPOSITION )) +
  geom_histogram(binwidth = 75, color = "black") 
```

```{r}
most_common_violation_disc <- table(chicago_violation_of_ward_15_3$VIOLATION.DESCRIPTION) %>%
sort(decreasing = TRUE)
most_common_violation_disc

chicago_violation_of_ward_15_4 <- filter(chicago_violation_of_ward_15_3, 
                                         VIOLATION.DESCRIPTION %in% c("Arrange for inspection of premises. (13-12-100)",
                                                                      "Post name, address, and telephone of owner, owner's agent for managing, controlling or collecting rents, and any other person managing or controlling building conspicuously where accessible or visible to public way. (13-12-030)",
                                                                      "Repair exterior wall. (13-196-010, 13-196-530 B)",
                                                                      "Repair or replace defective or missing members of porch system. (13-196-570)",
                                                                      "Replace broken, missing or defective window panes. (13-196-550 A)")
)


View(chicago_violation_of_ward_15_4)

unique_violation_descriptions <- unique(chicago_violation_of_ward_15_4$VIOLATION.DESCRIPTION)
print(paste("unique violation descriptions are:",unique_violation_descriptions))

shortened_names <- c("Inspection", "Owner Info", "Wall Fix", "Porch Repair", "Window Replace")


chicago_violation_of_ward_15_4$VIOLATION.DESCRIPTION <- factor(chicago_violation_of_ward_15_4$VIOLATION.DESCRIPTION,
                                                               levels = c("Arrange for inspection of premises. (13-12-100)",
                                                                          "Post name, address, and telephone of owner, owner's agent for managing, controlling or collecting rents, and any other person managing or controlling building conspicuously where accessible or visible to public way. (13-12-030)",
                                                                          "Repair exterior wall. (13-196-010, 13-196-530 B)",
                                                                          "Repair or replace defective or missing members of porch system. (13-196-570)",
                                                                          "Replace broken, missing or defective window panes. (13-196-550 A)"),
                                                               labels = shortened_names)

head(chicago_violation_of_ward_15_4$VIOLATION.DESCRIPTION)
```

```{r}
most_common_violation_disc <- table(chicago_violation_of_ward_15_4$VIOLATION.DESCRIPTION) %>%
  sort(decreasing = TRUE)
most_common_violation_disc

Avg_Imposed_fine_3<-chicago_violation_of_ward_15_4 %>%
  group_by(VIOLATION.DESCRIPTION) %>%
  summarise(AVG_IMPOSED_FINE = mean(IMPOSED.FINE , na.rm = TRUE)) %>%
  arrange(AVG_IMPOSED_FINE)

Avg_Imposed_fine_3$VIOLATION.DESCRIPTION <- reorder(Avg_Imposed_fine_3$VIOLATION.DESCRIPTION, Avg_Imposed_fine_3$AVG_IMPOSED_FINE, FUN = function(x) mean(x, na.rm = TRUE))


ggplot(Avg_Imposed_fine_3, aes(x = VIOLATION.DESCRIPTION, y = AVG_IMPOSED_FINE, fill = VIOLATION.DESCRIPTION)) +
  geom_bar(stat = "identity") +
  labs(title = "Imposed Fines",
       x = "Violation Description",
       y = "Average Imposed Fine") +
  scale_fill_manual(values = c("Inspection" = "blue", "Owner Info" = "green", "Wall Fix" = "orange", "Window Replace" = "purple", "Porch Repair" = "red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
# the plost shows that the average imposed fines for window replace is higher. 

#Violations related to window replacement might incur higher fines due to some reasons like:

#1. Safety concerns: Windows play a crucial role in providing ventilation, natural light, and security to a building. Any violation related to window replacement may pose safety risks, such as exposure to weather elements or compromised security, prompting authorities to impose larger fines to ensure compliance and prevent potential hazards.
#2.Regulatory standards: Building codes and regulations often mandate specific requirements for windows regarding size, material, insulation, and safety features. Violations related to window replacement may indicate non-compliance with these standards, which could lead to larger fines as a means to enforce adherence to regulatory requirements.
#3.Cost of rectification: Replacing windows can be a costly endeavor, especially if it involves multiple units or specialized materials. Fines for violations related to window replacement may reflect the estimated cost of rectifying the issue, considering factors such as labor, materials, and potential disruption to occupants.
#4. Impact on energy efficiency: Windows play a significant role in a building's energy efficiency by regulating heat gain, loss, and natural lighting. Violations that compromise the energy efficiency of windows, such as installing inefficient or non-compliant units, may result in larger fines to address the environmental impact and encourage compliance with energy efficiency standards.
#Overall, violations related to window replacement may attract higher fines due to the critical role of windows in building safety, regulatory requirements, the cost of rectification, and their impact on energy efficiency and occupant comfort.

```{r}
m <- leaflet() %>%
  addTiles() %>%
  addCircles(lng = chicago_violation_of_ward_15_4$LONGITUDE, lat = chicago_violation_of_ward_15_4$LATITUDE)
m



m1 <- leaflet() %>%
  addTiles() %>%
  addCircles(lng = chicago_violation_of_ward_15_4$LONGITUDE, lat = chicago_violation_of_ward_15_4$LATITUDE) %>%
  addProviderTiles(providers$Esri.WorldImagery)  
m1

```




